A Software Engineering Perspective on \\ OpenMP and PThreads

### 2017-08-09 sections/02-apis.tex
Minor revisions and spacing changes.

The C/C++ \textit{Pthreads} library controls thread execution in GNU/Linux by
creating and synchronizing threads. In this model, the execution begins in a
sequential region and threads must be explicitly started to begin
a parallel region.
Listing \ref{lst:listing-pthreads} presents a sample code using the
\textit{Pthreads} library. The program begins execution in a
single thread and until line 6 only one thread is running.

When \texttt{pthread\_create} is called a thread is created and the application
enters a parallel region with two simultaneous threads.  The main mechanism
available for synchronizing thread executions is the barrier. The
\textit{Pthreads} library implements barriers as calls to
\texttt{pthread\_join}. A thread that reaches a barrier will wait for other
threads. Line 11 shows an example of thread synchronization using a barrier.

\begin{lstlisting}[language=C, basicstyle=\ttfamily\scriptsize, numbers=left,
                   frame=no, showspaces=false, showstringspaces=false,
                   caption={\textit{Pthreads} Sample Program}, captionpos=b,
                   numberstyle=\tiny,
                   xleftmargin=0.5cm,
                   label=lst:listing-pthreads, keywords={%
                       DATATYPE, pthread_t, pthread_create,
                       pthread_join, task_function, NULL, int, main,
                       void, printf, return%
                       },
                   otherkeywords={::, \#pragma, \#include, <<<,>>>, \&, \*, +, -, /, [, ], >, <}
                   ]
int main(){
    pthread_t thread_id;
    DATATYPE *parameters = 10;
    int return_code;

    return_code = pthread_create(&thread_id, NULL,
                                 task_function,
                                 (void *) parameters);

    printf("Return Code: %d\n", return_code );
    pthread_join(thread_id , NULL);

    return 0;
}

void *task_function(void *data){
    DATATYPE *value;
    value = (DATATYPE *) data;
}
\end{lstlisting}

@@

Listing \ref{lst:listing-omp} presents an \textit{OpenMP} example. The
declaration of a parallel region with the \texttt{firstprivate} pragma on line
1 indicates the variable \texttt{aux\_dot} should be private to every thread
and initialized with the enclosing scope's value.  Line 3 will be executed by
only one thread, because of the \texttt{single} pragma.
The \texttt{for} in line 7 will be executed in parallel, with automatic work
division.  The pragma at line 12 defines a critical region which can only be
executed by a single thread at a time. The pragma at line 15 defines a
region that should only be executed by the master thread.

\begin{lstlisting}[language=C, basicstyle=\ttfamily\scriptsize, numbers=left,
                   frame=no, showspaces=false, showstringspaces=false,
                   caption={\textit{OpenMP} Sample Program}, captionpos=b,
                   numberstyle=\tiny,
                   xleftmargin=0.6cm,
                   label=lst:listing-omp, keywords={%
                       \#pragma,
                       omp, parallel, firstprivate,
                       single, omp_get_num_threads,
                       for, schedule, auto,
                       critical, master,
                       NULL, int, main,
                       void, printf, return%
                   },
                   otherkeywords={::, \#pragma, \#include, <<<,>>>, \&, \*, +, -, /, [, ], >, <}
                       ]
#pragma omp parallel firstprivate(aux_dot){
    #pragma omp single
    printf("Start of parallel region,
            number of threads: %d\n",
            omp_get_num_threads ());

>>>>

The C/C++ \textit{Pthreads} library controls thread execution in GNU/Linux by
creating and synchronizing threads. In this model, the execution begins in a
sequential region and threads must be explicitly started to begin a parallel
region.  Listing~\ref{lst:pthreads} presents a \textit{Pthreads}
example. The code shows the parallel computation of the dot product of two
arrays.

The program in Listing~\ref{lst:pthreads} runs in a single thread until
line 22, when \texttt{pthread\_create} is called.  Each call
in the loop initializes a new thread.
The parallelization in this example divides the arrays in chunks
and shares multiplications between threads.
One thread synchronization tool available in \textit{Pthreads} is the barrier,
implemented as calls to \texttt{pthread\_join}. A thread that reaches a barrier
waits for other threads. Line 27 shows an example of thread synchronization
using a barrier.
To avoid race conditions on the \texttt{dot} shared variable
the example uses \textit{mutexes} to create a critical
region, as seen in lines 42 and 44.

\begin{lstlisting}[language=C, basicstyle=\ttfamily\scriptsize, numbers=left,
                   frame=no, showspaces=false, showstringspaces=false,
                   caption={\textit{Pthreads} Sample Program}, captionpos=b,
                   numberstyle=\tiny,
                   xleftmargin=0.5cm,
                   label=lst:pthreads, keywords={%
                       DATATYPE, pthread_t, pthread_create,
                       pthread_join, task_function, NULL, int, main,
                       void, printf, return, pthread_mutex_t,
                       pthread_attr_t, pthread_attr_init,
                       MAX_THREADS, SIZE, char, struct, malloc,
                       MIN, pthread_mutex_lock, pthread_mutex_unlock,
                       pthread_exit%
                       },
                   otherkeywords={::, \#pragma, \#include, <<<,>>>, \&, \*, +, -, /, [, ], >, <}
                   ]
int A[SIZE], B[SIZE]; int dot = 0;
pthread_mutex_t lock;

int main(int argc, char *argv []){
  int num_threads, i, part_size;
  pthread_t tid [MAX_THREADS];
  pthread_attr_t attr;
  pthread_attr_init(&attr);
  /* Array initialization supressed */

  part_size = (SIZE + num_threads - 1) / num_threads;

  for (i = 0; i < num_threads; i++){
      struct arg_struct *args = \
              malloc(sizeof(struct arg_struct));

      args->id = i;
      args->i_part = (i * part_size);
      args->e_part = MIN(((i + 1) * (part_size)),
                         SIZE);

      pthread_create(&(tid[i]) ,&attr,
                     task_function,
                     (void *) args);
  }
  for (i = 0; i < num_threads; i++) {
      pthread_join(tid[i], NULL);
  }
  printf("Result: %d.\n", dot);
  return 0;
}

void *task_function (void *arguments){
  struct arg_struct *args = (arg_struct *) arguments;
  int i, partial_dot = 0;

  for(i = args->i_part; i < args->e_part; i++){
      partial_dot += A[i] * B[i];
  }

  /* Critical region */
  pthread_mutex_lock(&lock);
  dot += partial_dot;
  pthread_mutex_unlock(&lock);
  pthread_exit(0);
}
\end{lstlisting}

@@

Listing \ref{lst:omp} presents an \textit{OpenMP} example for the
same computation shown in Listing \ref{lst:pthreads}. The
declaration of a parallel region with the \texttt{firstprivate} pragma on line
1 indicates the variable \texttt{aux\_dot} should be private to every thread
and initialized with the enclosing scope's value.  Line 3 will be executed by
only one thread, because of the \texttt{single} pragma.
The loop in line 8 will be executed in parallel, with automatic work
division.  The pragma at line 12 defines a critical region which can only be
executed by a single thread at a time. The pragma at line 15 defines a
region that should only be executed by the master thread.

\begin{lstlisting}[language=C, basicstyle=\ttfamily\scriptsize, numbers=left,
                   frame=no, showspaces=false, showstringspaces=false,
                   caption={\textit{OpenMP} Sample Program}, captionpos=b,
                   numberstyle=\tiny,
                   xleftmargin=0.6cm,
                   label=lst:omp, keywords={%
                       \#pragma,
                       omp, parallel, firstprivate,
                       single, omp_get_num_threads,
                       for, schedule, auto,
                       critical, master,
                       NULL, int, main,
                       void, printf, return%
                   },
                   otherkeywords={::, \#pragma, \#include, <<<,>>>, \&, \*, +, -, /, [, ], >, <}
                       ]
#pragma omp parallel firstprivate(aux_dot){
  #pragma omp single
  printf("Start of parallel region,
          number of threads: %d\n",
          omp_get_num_threads ());
